Constructive forcing

### 2023-11-23 slides-herrsching2023.tex
More notes

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Maximal ideals}
  \jnote{1-}{
    The theorem on the slide is a generalization of a fact from undergraduate
    linear algebra: Over a field, no surjective matrix can have more rows than
    columns. (``Surjective'' here means that the induced linear map is
    surjective.)
  }

  \jnote{6-}{
    The slide presents a standard proof as offered by most textbooks on
    commutative algebra. The proof is quite efficient from a viewpoint of
    mathematical organization, as it quickly succeeds in reducing to the field
    situation. As such, it is short and memorable.

@@

    For this reason, the theorem and its classical proof are often used as case
    studies for tools and techniques aiming to extract constructive content
    from classical proofs. One such technique employs constructive forcing.
    The first(?) constructive proof, found directly without using extraction
    techniques, is laid out in a
    \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
  }

@@

    \item In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable forcing extension}\pause, and \\
    \emph{first-order consequences} of its existence there \good{do hold} in
    the base universe.
  \end{enumerate}
\end{frame}

\begin{frame}{Maximal ideals}
  \textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.

  \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
  have~$u, v \in A$ with
  \[
    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
    \quad\text{and}\quad
    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
  \]
  Hence
  $
    1 = (vy) (ux) = (uy) (vx) = 0
  $. \qed
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
\begin{frame}{Infinite data}
  \vspace*{-1em}
  \[ \astikznodetransparentlycircled{xm}{7}\!,
    \quad \astikznodetransparentlycircled{x0}{4}\!,
    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
    \quad \only<4->{\ldots} \]
  {\centering\begin{tikzpicture}[remember picture,overlay]
    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
    \path[draw=mypurple,thick]
      (t1)
      to [out=-90, in=180] (t3)
      to [out=0, in=-90] (t2);
  \end{tikzpicture}\par}
  \medskip
  \pause

>>>>

% \begin{document}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Maximal ideals}
  \jnote{1-7}{
    The theorem on the slide is a generalization of a fact from undergraduate
    linear algebra: Over a field, no surjective matrix can have more rows than
    columns. (``Surjective'' here means that the induced linear map is
    surjective.)
  }

  \jnote{6-7}{
    The slide presents a standard proof as offered by most textbooks on
    commutative algebra. The proof is quite efficient from a viewpoint of
    mathematical organization, as it quickly succeeds in reducing to the field
    situation. As such, it is short and memorable.

@@

    For this reason, the theorem and its classical proof are often used as case
    studies for tools and techniques aiming to extract constructive content
    from classical proofs. One such technique employs constructive forcing.
    The first(?)\@ constructive proof, found directly without using extraction
    techniques, is laid out in a
    \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
  }

  \jnote{11}{
    The iterative construction in the countable case without decidability
    assumptions is due to Krivine; it was later clarified by Berardi and
    Valentini. It is a parlor trick, resulting in a subset which formally
    verifies the axioms for a maximal ideal but without carrying out any actual
    work. Indeed, the resulting ideal will in general not be a detachable
    subset of the ring.

    Surprisingly, there is still computational content in this construction, as
    explored in \fixedhref{https://arxiv.org/abs/2207.03873}{this joint paper
    with Peter Schuster}; one interpretation of our observation is that classical
    proofs don't ``really'' require a maximal ideal; they just use that notion
    for structuring hidden computations.
  }

  \jnote{13-}{
    In a suitable forcing extension, the ring appears countable. Hence we can
    carry out the iterative maximal ideal construction there. The resulting ideal
    will not be part of the base universe (instead, from the point of view of
    the base universe we will just have constructed a certain sheaf of ideals on a
    certain pointfree space), but bounded first-order consequences of its
    existence still pass down to the base.
  }

@@

    \item In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable forcing extension}\pause, and \\
    \emph{bounded first-order consequences} of its existence there \good{do hold} in
    the base universe.
  \end{enumerate}
\end{frame}

\begin{frame}{Maximal ideals}
  \jnote{1}{
    Unwinding all the definitions from constructive forcing and from the
    iterative maximal ideal construction, and eliminating the application
    of~\textsc{lem} from the classical proof presented before, we mechanically
    arrive at the constructive direct proof presented on the slide.
  }

  \textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.

  \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
  we have~$u, v \in A$ with
  \[
    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
    \quad\text{and}\quad
    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
  \]
  Hence
  $
    1 = (vy) (ux) = (uy) (vx) = 0
  $. \qed
\end{frame}}

% \begin{document}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
\begin{frame}{Infinite data}
  \jnote{5-}{
    The presented proof rests on the law of excluded middle and hence cannot
    immediately be interpreted as a program for finding suitable indices~$i < j$.
    However, constructive proofs are also possible (for instance by
    induction on the value of a given term of the sequence), and furthermore
    constructive proofs can be extracted from the presented classical proof.
  }

  \jnote{7-}{
    The class of well preorders is stable under cartesian products, lists and
    trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
    respectively. However, in their naive formulations, these are merely
    theorems of classical mathematics. For general constructive results, the
    definition of ``well'' needs to be improved.

    Classical texts often employ needless negations in their definitions; this
    is an XXX
  }

  \vspace*{-1em}
  \[ \astikznodetransparentlycircled{xm}{7}\!,
    \quad \astikznodetransparentlycircled{x0}{4}\!,
    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
    \quad \only<4->{\ldots} \]
  {\centering\begin{tikzpicture}[remember picture,overlay]
    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
    \path[draw=mypurple,thick]
      (t1)
      to [out=-90, in=180] (t3)
      to [out=0, in=-90] (t2);
  \end{tikzpicture}\par}
  \medskip
  \pause
